Brief notes on erosion

### 2017-10-14 docs/TexNotes/Notes.tex
highlighting in Saverio's paper?
Took debugging print lines out of thetalne.jl

@@

\section{Shear stress and erosion law}
The deformation, or rate-of-strain, tensor is given by
\begin{equation}
\bvec{E} = \frac{1}{2} \left( \grad \bu + \grad \bu^T \right)
\end{equation}
where $\bu = (u,v)$ is the velocity field. The stress tensor is given by
\begin{equation}
\bvec{\sigma} = -p \bvec{I} + 2 \mu \bvec{E} = -p \bvec{I} + \mu \left( \grad \bu + \grad \bu^T \right)
\end{equation}
The surface traction is given by
\begin{equation}
\bvec{f} = \bvec{\sigma} \cdot \nhat 	\qquad \onBdry
\end{equation}
where $\nhat$ is the (outward) unit normal and $\bdry$ is the solid boundary. The shear stress $\tau$ is defined as the tangential component of the traction, i.e.~
\begin{equation}
\label{stressdef}
\tau = \bvec{f} \cdot \shat 
= \mu \left( \grad \bu + \grad \bu^T \right) \cdot \nhat \cdot \shat	\qquad \onBdry
\end{equation}
where $\shat$ is the unit tangent vector. It turns out that the shear stress is simply given by the surface vorticity
\begin{equation}
\tau = \mu \left( u_y - v_x \right) \qquad \onBdry
\end{equation}
which can be shown by a direct calculation, using the no-slip boundary conditions (which implies that $\pdi{u}{s} = \pdi{v}{s} = 0$ on $\bdry$).
The interface velocity of the eroding body is linearly proportional to the absolute shear stress
\begin{equation}
V_n = -C \atau
\end{equation}
where $C$ is a constant.

@@

\section{Scaling law and interface smoothing}
For 3D Stokes flow, the shear stress scales like
\begin{equation}
\tau \sim \mu V/L
\end{equation}
where $V$ and $L$ are a characteristic velocity and length scale respectively. To evolve the interface in our numerical method, we will probably need to introduce some smoothing to mitigate high frequencies. A common approach is to introduce a small curvature term into the interface velocity, sometimes called the method of vanishing viscosity. The most basic implementation would be
\begin{equation}
V_n = - C \atau - \eps \kappa
\end{equation}
where $\eps$ is a small parameter and $\kappa$ is the surface curvature. However, we have two additional preferences
\begin{enumerate}
\item As the body changes size, we would like the influence of the curvature term relative to the shear stress to remain the same.
\item We prefer the curvature term not to change the size of the body.
\end{enumerate}
Fortunately, the first requirement is already satisfied since both $\tau$ and $\kappa$ scale like $1/L$. To satisfy the second requirement, we replace $\kappa$ with this same quantity minus its mean. To allow for some generality, we allow the curvature term by to be scaled by a power of $L$, giving
\begin{equation}
V_n = - C \atau - \eps L^{\beta} \left(\kappa - \frac{2 \pi}{L} \right)
\end{equation}
Thus, for the present case of Stokes flow $\beta=0$, but for the previous case of high-Reynolds-number flow $\beta=1/2$.

@@

To evolve the interface numerically, we use the $\theta$-$L$ method. Rather than re-derive the formulas, I will use the convention from Shelley 1994, with the curve parameterized in the CCW direction and the normal vector pointing {\em inward}. With this convention, the normal interface velocity is the negative of what it was previously, giving
\begin{equation}
\label{Vn1}
V_n = C \atau + \eps L^{\beta} \left(\kappa - \frac{2 \pi}{L} \right)
\end{equation}
Recall that for Stokes flow, $\beta = 0$.

Now, the interface $\X = (x,y)$ evolves according to
\begin{equation}
\pd{\bvec{X}}{t} = V_n \nhat + V_s \shat
\end{equation}
where $\nhat = (-y_s, x_s)$ is the inward pointing normal vector and $\shat = (x_s, y_s)$ is the tangential vector (running in the CCW direction). Here, we have introduced a tangential velocity $V_s$ that will be chosen to keep the curve in the equal-arclength parameterization. We will frequently use a normalized arclength $\alpha = s/L$, where $L$ is the total length of the curve.

We introduce the tangent angle $\theta$, defined by
\begin{equation}
(x_s, y_s) = (\cos \theta, \sin \theta)
\end{equation}
so that we have
\begin{equation}
\tan \theta = y_s/x_s = y_\alpha / x_\alpha
\end{equation}
Remember the identity that $x_s^2 + y_s^2 = 1$ as this is very useful.
The curvature is given by
\begin{equation}
\kappa = x_s y_{ss} - x_{ss} y_s = \pd{\theta}{s} = L^{-1} \pd{\theta}{\alpha}
\end{equation}
We are using the standard convention, where a convex shape, like a circle, has positive curvature. The normal velocity now can be written as
\begin{equation}
\label{Vn2}
V_n = C \atau + \eps L^{\beta-1} \left(\pd{\theta}{\alpha} - 2 \pi \right)
\end{equation}

We now discuss how to choose $V_s$ to remain in the equal-arclength frame of reference. We will use the term
\begin{equation}
\gamma = \pd{\theta}{\alpha} V_n
\end{equation}
Then the necessary tangential velocity is determined by
\begin{equation}
\td{V_s}{\alpha} = \gamma - \mean{\gamma}
\end{equation}
where $\mean{}$ indicates the mean value.

We now discuss how to evolve $\theta$ and $L$ forward in time according to the interface velocity given above. The resulting evolution equations are
\begin{align}
& \td{L}{t} = - \mean{\gamma} \\
& \pd{\theta}{t} = L^{-1} \left( \pd{V_n}{\alpha} + \pd{\theta}{\alpha} V_s \right)
\end{align}
Then, using the above $V_n$, the equation for $\theta$ simplifies to
\begin{equation}
\label{thetaEvolve}
\pd{\theta}{t} = \eps L^{\beta-2} \ppd{\theta}{\alpha} + L^{-1} \left( C \pd{\atau}{\alpha} + \pd{\theta}{\alpha} V_s \right)
\end{equation}

We now discuss the numerical methods used to solve the above evolution equations. For $L$, we use an explicit method
\begin{equation}
L^{(n+1)} = L^{(n)} + \frac{\Dt}{2} \left( 3 M^{(n)} - M^{(n-1)} \right)
\end{equation}
where
\begin{equation}
M = \td{L}{t} = -\mean{\gamma}
\end{equation}
For $\theta$, we use a spectral method, giving
\begin{equation}
\td{\thh_k}{t} + 4 \pi^2 k^2 \eps L^{\beta-2} \thh_k = \hat{N}_k
\end{equation}
where the nonlinear terms are grouped together as
\begin{equation}
N = L^{-1} \left( C \pd{\atau}{\alpha} + \pd{\theta}{\alpha} V_s \right)
\end{equation}
We evolve with the integrating-factor method. When recast in terms of Gaussian filters, this gives
\begin{equation}
\label{intfact}
\thh^{(n+1)} = \GG(\theta^{(n)}, \sigma_1) 
+ \frac{\Dt}{2} \left( 3 \GG(N^{(n)}, \sigma_1) - \GG(N^{(n-1)}), \sigma_2 \right)
\end{equation}
where
\begin{align}
& \sigma_1 = 2 \pi \sqrt{\eps \Dt} \sqrt{(L^{n})^{\beta-2} + (L^{n+1})^{\beta-2} } 	\\
& \sigma_2 = 2 \pi \sqrt{\eps \Dt} \sqrt{(L^{n-1})^{\beta-2} + 2 (L^{n})^{\beta-2} + (L^{n+1})^{\beta-2} }
\end{align}

\subsection{Different scaling of the stress}
Some numerical tests reveal that for 2D Stokes flow the stress scales like $\tau \sim 1/(L \log L)$, instead of the naive $1/L$. This is because we are in a 2D bounded domain (Stokes paradox type stuff). So we need to modify the curvature driven flow to have the same scaling. To this end, modify Eq.~(\ref{Vn1}) to read
\begin{equation}
V_n = \atau + \eps f(L) \left(\kappa - \frac{2 \pi}{L} \right)
\end{equation}
For 2D Stokes flow, we have 
\begin{equation}
f(L) = \frac{1}{\log (W/L)}
\end{equation}
where $W$ is the lengthscale of the 3rd dimension (e.g. the length of a long cylinder). The 2D approximation requires $W \gg L$, and in the code I chose $W = 2 \pi$. Carrying through this change, Eq.~(\ref{Vn2}) becomes
\begin{equation}
V_n = \atau + \eps L^{-1} f(L) \left(\pd{\theta}{\alpha} - 2 \pi \right)
\end{equation}
and Eq.~(\ref{thetaEvolve}) becomes
\begin{equation}
\pd{\theta}{t} = \eps L^{-2} f(L) \ppd{\theta}{\alpha} + L^{-1} \left( \pd{\atau}{\alpha} + \pd{\theta}{\alpha} V_s \right)
\end{equation}
Then to evolve $\theta$ spectrally, we use
\begin{equation}
\td{\thh_k}{t} + 4 \pi^2 k^2 \eps L^{-2} f(L)  \thh_k = \hat{N}_k
\end{equation}
where the nonlinear terms are grouped together as
\begin{equation}
N = L^{-1} \left( \pd{\atau}{\alpha} + \pd{\theta}{\alpha} V_s \right)
\end{equation}
We evolve with the integrating-factor method. When recast in terms of Gaussian filters, this gives
\begin{equation}
\label{intfact}
\thh^{(n+1)} = \GG(\theta^{(n)}, \sigma_1) 
+ \frac{\Dt}{2} \left( 3 \GG(N^{(n)}, \sigma_1) - \GG(N^{(n-1)}), \sigma_2 \right)
\end{equation}
where
\begin{align}
& \sigma_1 = 2 \pi \sqrt{\eps \Dt} \sqrt{g(L^{n}) + g(L^{n+1}) } 	\\
& \sigma_2 = 2 \pi \sqrt{\eps \Dt} \sqrt{g(L^{n-1}) + 2 g(L^{n}) + g(L^{n+1}) }
\end{align}
where $g(L) = L^{-2} f(L)$.

\subsection{The vanishing rate of the area}
The rate of area loss is given by
\begin{equation}
\dot{A} = -C \abs{\tau} L
\end{equation}
We then insert the scaling $\abs{\tau} \sim -1/(L \log L)$ and simplify to get
\begin{equation}
\dot{A} = C/\log{A}
\end{equation}
which we solve to get
\begin{equation}
A (1 - \log A) = C (t_f - t)
\end{equation}
In particular, I believe the signs are correct here.
BUT, I am not so sure of this calculation anymore, now that $W$ is included.

@@

The calculation is based on looking at Eq.~(\ref{intfact}), where $\sigma_1 \sim 2 \pi \sqrt{\eps \Delta t}$. For the largest wavenumbers, $k_{max} \sim N/2$, we want the smoothing to beat the nonlinear term (so that overall, these modes decay). So we need $e^{-\pi^2 \eps \Delta t N^2 / 2} < 1 - C\Delta t$, which gives $\eps > 2C/(\pi^2 N^2)$, independent of $\Delta t$. If $C=1$, this gives roughly $\eps > 0.2/ N^2$. However, this analysis did not take into account how the spectrum of the nonlinear term decays (or maybe it even grows???)

>>>>

% FROM METHODS TEX
\newcommand{\xx}{{\mathbf{x}}}
\renewcommand{\ss}{{\mathbf{s}}}
\newcommand{\nn}{{\mathbf{n}}}
% Derivatives
\newcommand{\pderiv}[2]{\frac{\partial #1}{\partial #2}}
\newcommand{\tderiv}[2]{\frac{d #1}{d #2}}
% Nick's commands
\newcommand{\thL}{$\theta$--$L$}
\newcommand{\Vn}{V_n}
\newcommand{\Vs}{V_s}
\newcommand{\thalpha}{\pderiv{\theta}{\alpha}}
\newcommand{\elfun}{\zeta}
\newcommand{\thhat}{\hat{\theta}}

@@

\section{Shear stress and erosion law}
The deformation, or rate-of-strain, tensor is given by
\begin{equation}
\bvec{E} = \frac{1}{2} \left( \grad \bu + \grad \bu^T \right)
\end{equation}
where $\bu = (u,v)$ is the velocity field. The stress tensor is given by
\begin{equation}
\bvec{\sigma} = -p \bvec{I} + 2 \mu \bvec{E} = -p \bvec{I} + \mu \left( \grad \bu + \grad \bu^T \right)
\end{equation}
The surface traction is given by
\begin{equation}
\bvec{f} = \bvec{\sigma} \cdot \nhat 	\qquad \onBdry
\end{equation}
where $\nhat$ is the (outward) unit normal and $\bdry$ is the solid boundary. The shear stress $\tau$ is defined as the tangential component of the traction, i.e.~
\begin{equation}
\label{stressdef}
\tau = \bvec{f} \cdot \shat 
= \mu \left( \grad \bu + \grad \bu^T \right) \cdot \nhat \cdot \shat	\qquad \onBdry
\end{equation}
where $\shat$ is the unit tangent vector. It turns out that the shear stress is simply given by the surface vorticity
\begin{equation}
\tau = \mu \left( u_y - v_x \right) \qquad \onBdry
\end{equation}
which can be shown by a direct calculation, using the no-slip boundary conditions (which implies that $\pdi{u}{s} = \pdi{v}{s} = 0$ on $\bdry$).
The interface velocity of the eroding body is linearly proportional to the absolute shear stress
\begin{equation}
V_n = -C \atau
\end{equation}
where $C$ is a constant. In what follows, we will scale time on an erosion timescale (non-dimensionalize) and thus set $C=1$.

@@

\section{Scaling law and interface smoothing}
For 3D Stokes flow, the shear stress scales like
\begin{equation}
\tau \sim \mu V/L
\end{equation}
where $V$ and $L$ are a characteristic velocity and length scale respectively.  However, for 2D Stokes flow we have Stokes paradox and slender body effects. The analogous scaling of the stress is
\begin{equation}
\tau \sim \frac{\mu V}{L \log(W/L)}
\end{equation}
where $W$ is the width in the third dimension. In the following, we will take $W = 2\pi$, since a body cannot have a length greater than $2\pi$ and still fit inside the domain.

\subsection{The vanishing rate of the area}
The rate of area loss is given by
\begin{equation}
\dot{A} = - \abs{\tau} L
\end{equation}
We then insert the scaling $\abs{\tau} \sim -1/(L \log L)$ and simplify to get
\begin{equation}
\dot{A} = C/\log{A}
\end{equation}
which we solve to get
\begin{equation}
A (1 - \log A) = C (t_f - t)
\end{equation}
In particular, I believe the signs are correct here.
BUT, I am not so sure of this calculation anymore, now that $W$ is included.

@@

To handle boundary evolution, we use the {\thL} method, which offers certain advantages in the ability to stabilize fluid-structure interaction problems. Throughout this section, we use the following conventions: the bodies are parametrized in the CCW direction and the normal vector points into the bodies (our out of the fluid).

In order to regularize boundary evolution, we modify the normal velocity as follows
\begin{equation}
\label{eqn:Vn}
\Vn = \atau + \frac{\eps}{\log \left(2\pi/L \right)} \left(\kappa - \frac{2 \pi}{L} \right)
\end{equation}
Here, $\kappa$ is the curvature, $\eps$ is a small parameter, and $L$ is the total arc length of the body. The second term on the right acts to smooth bodies that have high-frequency oscillations. This is sometimes called the method of vanishing viscosity. Since we have subtracted the mean curvature, $2\pi/L$, this term is area-preserving, so that the only material loss is due to the shear stress. We chose the prefactor $\log^{-1} \left(2\pi/L \right)$, so that, for 2D Stokes flow, this term scales the same as shear stress as $L \to 0$. Thus, the relative strength of the smoothing is always the same.

Note: The curvature is given by
\begin{equation}
\kappa = x_s y_{ss} - x_{ss} y_s = \pd{\theta}{s} = L^{-1} \pd{\theta}{\alpha}
\end{equation}
We are using the standard convention, where a convex shape, like a circle, has positive curvature.

We now introduce the tangent angle, $\theta$, defined through the relation
\begin{equation}
\left( \pderiv{x}{s}, \pderiv{y}{s} \right) = \left(\cos \theta, \sin \theta \right)
\end{equation}
where $s$ is arc length. We also introduce the normalized arc length $\alpha = s / L$.

Note: We have this relation
\begin{equation}
\tan \theta = y_s/x_s = y_\alpha / x_\alpha
\end{equation}
and remember the identity that $x_s^2 + y_s^2 = 1$ as this is very useful.

In terms of these new variables, the curvature can be expressed as
\begin{equation}
\kappa = \frac{1}{L} \pderiv{\theta}{\alpha}
\end{equation}
Substitution into Eq.~(\ref{eqn:Vn}) gives
\begin{equation}
\label{eqn:Vn2}
\Vn = \atau + \frac{\eps}{L \log \left(2\pi/L \right)}  \left(\pderiv{\theta}{\alpha} - 2 \pi \right)
\end{equation}
The main idea is to introduce an artificial tangential velocity that keeps collocation points equally spaced with respect to arc length; i.e.~the grid of $\alpha$ should always remain equally spaced. The tangential velocity that does the trick is determined by Shelley 1994 as
\begin{equation}
\pderiv{\Vs}{\alpha} = \thalpha \Vn - \mean{\thalpha \Vn}
\end{equation}
The interface evolution is then given by
\begin{equation}
\dot{\xx}(t) = \Vn \nn + \Vs \ss.
\end{equation}

Note: with the above conventions, we have $\nn = (-y_s, x_s)$ as the inward pointing normal vector and $\ss = (x_s, y_s)$ as the tangential vector (running in the CCW direction).

Upon transforming to the variables $\theta(\alpha,t)$ and $L(t)$, we obtain the evolution equations
\begin{align}
& \tderiv{L}{t} = - \mean{\thalpha \Vn} \\
& \pderiv{\theta}{t} = \frac{1}{L} \left( \pderiv{\Vn}{\alpha} + \thalpha \Vs \right)
\end{align}
Inserting Eq.~(\ref{eqn:Vn2}) for the normal velocity gives
\begin{equation}
\pderiv{\theta}{t} = \frac{\eps}{L^{2} \log \left(2\pi/L \right)} \ppd{\theta}{\alpha} 
+ \frac{1}{L} \left( \pderiv{\atau}{\alpha} + \thalpha \Vs \right)
\end{equation}
For simplicity, we introduce the variables
\begin{align}
& M = - \mean{\thalpha \Vn} \\
& N = \frac{1}{L} \left( \pderiv{\atau}{\alpha} + \thalpha \Vs \right) \\
& \elfun = \frac{1}{L^{2} \log \left(2\pi/L \right)}
\end{align}
Then the evolution equations can be expressed as
\begin{align}
& \tderiv{L}{t} = M \\
& \pderiv{\theta}{t} = \eps \elfun \ppd{\theta}{\alpha} + N
\end{align}
We treat $\theta$ spectrally
\begin{equation}
\theta(\alpha,t) = \sum_{k = -\infty}^{\infty} \thhat_k e^{2\pi i k \alpha}
\end{equation}
Then we get
\begin{equation}
\tderiv{\thhat_k}{t} +  4 \eps \pi^2 k^2  \elfun \thhat_k = \hat{N}_k
\end{equation}
We introduce the integrating factor
\begin{equation}
\mu = \exp \left( 4 \eps \pi^2 k^2 \int \elfun dt \right)
\end{equation}
so that the solution can be represented as
\begin{align}
\thhat_k(t) & = \frac{1}{\mu} \int^t \mu \hat{N}_k dt'
= \thhat_k(0) + \int_0^t \exp \left( -4 \eps \pi^2 k^2 \int_{t'}^t \elfun dt'' \right) \hat{N}_k \, dt'
% LONGER FORM
%& = \exp \left( -4 \eps \pi^2 k^2 \int_0^t \elfun(L) ds \right) \, 
%\int_0^t \exp \left( 4 \eps \pi^2 k^2 \int_0^s \elfun(L) ds' \right) \hat{N}_k ds \\
\end{align}
where $t'$ and $t''$ are simply dummy variables for integration. We can get the exact relationship
\begin{equation}
\thhat_k(t + \Delta t) = \thhat_k(t) 
+ \int_t^{t+\Delta t} \exp \left( -4 \eps \pi^2 k^2 \int_{t'}^{t+\Delta t} \elfun dt'' \right) \hat{N}_k \, dt'
\end{equation}

@@

The calculation is based on looking at Eq.~?, where $\sigma_1 \sim 2 \pi \sqrt{\eps \Delta t}$. For the largest wavenumbers, $k_{max} \sim N/2$, we want the smoothing to beat the nonlinear term (so that overall, these modes decay). So we need $e^{-\pi^2 \eps \Delta t N^2 / 2} < 1 - C\Delta t$, which gives $\eps > 2C/(\pi^2 N^2)$, independent of $\Delta t$. If $C=1$, this gives roughly $\eps > 0.2/ N^2$. However, this analysis did not take into account how the spectrum of the nonlinear term decays (or maybe it even grows???)
